Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-03-31 Paper/paper.tex
updated Mchirp error figure

% #################
\begin{figure*}
\centering 
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR}.pdf
\caption{We show here (Log of) the ratio of the systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering    
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\caption{These figures are similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
NS tidal effects on GW emission can become the dominant source of error. However,
for quieter signals with $\rho\leq 30$, we find that not including NS matter 
effects in parameter estimation templates has minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
\includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{These figures show the ratio of the systematic and statistical
measurement errors for BH spins, for all of our injected cases. The left
panel corresponds to $1-\sigma$ statistical confidence level, while the right panel
corresponds to $90\%$ confidence level. We note that below $\rho\approx 30$, ignoring
tidal effects in templates introduces systematic effects that remain subdominant
to the statistical measurement uncertainties.
}
\label{fig:TNT_BHspinBias_vs_SNR_q23}
\end{figure*}
% 
% 
Advanced LIGO searches and parameter estimation efforts aimed at binaries 
containing NS and stellar-mass BHs are poised to use waveform models that do not
include the effects of the NS tidal deformability~\cite{Canton:2014ena}.
While this is not expected to
be the dominant source of error at low SNRs, it will likely introduce a
systematic bias in the physical parameters that we recover from GW observations.
In this section, we study the impact of the same for binaries where the BH spins
are aligned to the orbital angular momentum.

@@

We want to distribute signals in effective distance $\deff$, which is a combination of 
luminosity distance, inclination angle of the binary's orbital angular momentum
with respect to the detector, and its various sky angles. The signal strength 
scales inversely with $\deff$, if it comprises only of the dominant $(2,2)$ mode, allowing
for various angle-dependent factors to scale out beside the luminosity distance.
% 
In order to consider multiple events, lets imagine a population uniformly distributed
in effective volume, i.e. within a sphere of radius $\deff^\mathrm{max}$. The 
radius of this sphere is set by the lowest SNR that is distinguishable from noise
by LIGO searches. Now, dividie the sphere into $I$ shells of equal width. The radius
of the $i$-th sphere would then be $D_i = \deff^\mathrm{max} (i - 1/2)/I$. The 
measurement uncertainty in $\lambdans$ scales inversely with the SNR, and hence
directly with the effective distance to the source. Therefore, if we have a 
measurement error $\sigma_0$ for a source located at $\deff = D_0$, the same error
for the same source located within the $i$-th shell would be 
$$
\sigma_i = \sigma_0 \frac{D_i}{D_0}.
$$
Independent measurements of $\lambdans$ for identical sources at different distances
would have their combined error given by
\begin{equation}\label{eq:1oversigma}
\frac{1}{\sigma^2} = \sum_{i=1}^I \frac{N_i}{\sigma_i^2} = \left(\frac{D_0}{\sigma_0}\right)^2 \sum_{i=1}^I\frac{N_i}{D_i^2},
\end{equation}
where $N_i$ is the number of sources detected in the $i$-th shell, and is a random
variable, with its probability proportional to the volume of the shell, i.e.
$$
p(N_i) \propto \frac{V_i}{V_\mathrm{total}} \propto \dfrac{\left(\deff^\mathrm{max} \frac{i}{I}\right)^3 - \left(\deff^\mathrm{max} \frac{i-1}{I}\right)^3}{(\deff^\mathrm{max})^3} \propto \frac{1}{I^3} [i^3 - (i-1)^3].
$$
Therefore, the expected number of detections in the $i$-th shell would be
$$
\langle N_i\rangle = \frac{N}{I^3} [i^3 - (i-1)^3],
$$
where $N=\sum_{i=1}^I N_i$ is the total number of NSBH detections. $N$ is expected
to be Poisson distributed around the mean detection rate
$\mathcal{R}\equiv\langle N\rangle$. Let us assume that we have $\mathcal{R}_d$ detections
per year per unit effective volume, where $\mathcal{R}_d$ can vary from $0.6-1000$ per
$\mathrm{Gpc}^3$ per year~\cite{Abadie:2010cfa}. Then 
\begin{equation}
\langle N\rangle = \int_0^{\deff^\mathrm{max}} 4\pi D^2 \mathcal{R}_d \D D = \frac{4\pi}{3} \mathcal{R}_d (\deff^\mathrm{max})^3,
\end{equation}
and
\begin{eqnarray}
 \langle \frac{1}{\sigma^2}\rangle &=& \left(\frac{D_0}{\sigma_0}\right)^2 \int_0^{\deff^\mathrm{max}} \frac{4\pi\mathcal{R}_d D^2 }{D^2}\D D\\
 &=& \left(\frac{D_0}{\sigma_0}\right)^2 4\pi \mathcal{R}_d \deff^\mathrm{max},
\end{eqnarray}
where in the previous equation we have converted the summation in Eq.~\ref{eq:1oversigma} 
to an integral. So the root-mean-square averaged measurement error from $N$ sources 
distributed uniformly in effective volume within a sphere of radius $\deff^\mathrm{max}$
would be
\begin{equation}\label{eq:rmsSigmaIdenticalSources}
 \sigma_{avg} \equiv \frac{1}{\sqrt{\langle1/\sigma^{2}\rangle}} = \frac{\sigma_0}{D_0} \deff^\mathrm{max} \frac{1}{3\mathcal{R}_d}.
\end{equation}
Let us analyze the implications of Eq.~\ref{eq:rmsSigmaIdenticalSources}. It is
straightforward to deduce that the same measurement certainty as afforded by a single
observation with a high SNR $\alpha$ can be obtained from $N = \alpha^2/300$ observations
uniformly distributed in effective volume with SNRs $\geq 10$. E.g., to get to the level 
of certainty afforded at SNR$=70$, we would need $49/3\approx 16-17$ realistic detections.

>>>>

% #################
\begin{figure*}
\centering 
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
\includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR}
\caption{We show here (Log of) the ratio of the systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering    
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\caption{These figures are similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
NS tidal effects on GW emission can become the dominant source of error. However,
for quieter signals with $\rho\leq 30$, we find that not including NS matter 
effects in parameter estimation templates has minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
\includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{These figures show the ratio of the systematic and statistical
measurement errors for BH spins, for all of our injected cases. The left
panel corresponds to $1-\sigma$ statistical confidence level, while the right panel
corresponds to $90\%$ confidence level. We note that below $\rho\approx 30$, ignoring
tidal effects in templates introduces systematic effects that remain subdominant
to the statistical measurement uncertainties.
}
\label{fig:TNT_BHspinBias_vs_SNR_q23}
\end{figure*}
% 
% 
Advanced LIGO searches and parameter estimation efforts aimed at binaries 
containing NS and stellar-mass BHs are poised to use waveform models that do not
include the effects of the NS tidal deformability~\cite{Canton:2014ena}.
While this is not expected to
be the dominant source of error at low SNRs, it will likely introduce a
systematic bias in the physical parameters that we recover from GW observations.
In this section, we study the impact of the same for binaries where the BH spins
are aligned to the orbital angular momentum.

@@

We want to distribute signals in effective distance $\deff$, which is a combination of 
luminosity distance, inclination angle of the binary's orbital angular momentum
with respect to the detector, and its various sky angles. The signal strength 
scales inversely with $\deff$, if it comprises only of the dominant $(2,2)$ mode, allowing
for various angle-dependent factors to scale out beside the luminosity distance.
% 
In order to consider multiple events, lets imagine a population uniformly distributed
in effective volume, i.e. within a sphere of radius $\deff^\mathrm{max}$. The 
radius of this sphere is set by the lowest SNR that is distinguishable from noise
by LIGO searches. Now, dividie the sphere into $I$ shells of equal width. The radius
of the $i$-th sphere would then be $D_i = \deff^\mathrm{max} (i - 1/2)/I$. The 
measurement uncertainty in $\lambdans$ scales inversely with the SNR, and hence
directly with the effective distance to the source. Therefore, if we have a 
measurement error $\sigma_0$ for a source located at $\deff = D_0$, the same error
for the same source located within the $i$-th shell would be 
$$
\sigma_i = \sigma_0 \frac{D_i}{D_0}.
$$
Independent measurements of $\lambdans$ for identical sources at different distances
would have their combined error given by
\begin{equation}\label{eq:1oversigma}
\frac{1}{\sigma^2} = \sum_{i=1}^I \frac{N_i}{\sigma_i^2} = \left(\frac{D_0}{\sigma_0}\right)^2 \sum_{i=1}^I\frac{N_i}{D_i^2},
\end{equation}
where $N_i$ is the number of sources detected in the $i$-th shell, and is a random
variable, with its probability proportional to the volume of the shell, i.e.
$$
p(N_i) \propto \frac{V_i}{V_\mathrm{total}} \propto \dfrac{\left(\deff^\mathrm{max} \frac{i}{I}\right)^3 - \left(\deff^\mathrm{max} \frac{i-1}{I}\right)^3}{(\deff^\mathrm{max})^3} \propto \frac{1}{I^3} [i^3 - (i-1)^3].
$$
Therefore, the expected number of detections in the $i$-th shell would be
$$
\langle N_i\rangle = \frac{N}{I^3} [i^3 - (i-1)^3],
$$
where $N=\sum_{i=1}^I N_i$ is the total number of NSBH detections. $N$ is expected
to be Poisson distributed around the mean detection rate
$\mathcal{R}\equiv\langle N\rangle$, where $\mathcal{R}$ can vary from $0.6-1000$ per
$\mathrm{Gpc}^3$ per year~\cite{Abadie:2010cfa}. If there are $n$ resulting detections
a year per unit effective volume, then
\begin{equation}
\langle N\rangle = \int_0^{\deff^\mathrm{max}} 4\pi n D^2 \D D = \frac{4\pi}{3} n (\deff^\mathrm{max})^3,
\end{equation}
and
\begin{eqnarray}
 \langle \frac{1}{\sigma^2}\rangle &=& \left(\frac{D_0}{\sigma_0}\right)^2 \int_0^{\deff^\mathrm{max}} \frac{4\pi n D^2 }{D^2}\D D\\
 &=& \left(\frac{D_0}{\sigma_0}\right)^2 4\pi n \deff^\mathrm{max},
\end{eqnarray}
where in the previous equation we have converted the summation in Eq.~\ref{eq:1oversigma} 
to an integral. So the root-mean-square averaged measurement error from $N$ sources 
distributed uniformly in effective volume within a sphere of radius $\deff^\mathrm{max}$
would be
\begin{equation}\label{eq:rmsSigmaIdenticalSources}
 \sigma_{avg} \equiv \frac{1}{\sqrt{\langle1/\sigma^{2}\rangle}} = \frac{\sigma_0}{D_0} \deff^\mathrm{max} \frac{1}{\sqrt{3\langle N\rangle}}.
\end{equation}
Let us analyze the implications of Eq.~\ref{eq:rmsSigmaIdenticalSources}. It is
straightforward to deduce that the same measurement certainty as afforded by a single
observation with a high SNR $\alpha$ can be obtained from $N = \alpha^2/300$ observations
uniformly distributed in effective volume with SNRs $\geq 10$. E.g., to get to the level 
of certainty afforded at SNR$=70$, we would need $49/3\approx 16-17$ realistic detections.
